# Pattern Mining
## Library

```{r}
# DAL ToolBox
# version 1.0.77


library(daltoolbox)

library(arules)
library(arulesViz)
library(arulesSequences)
```

```{r}
data(AdultUCI)
dim(AdultUCI)
head(AdultUCI)
```

## Removing attributes

```{r}
AdultUCI$fnlwgt <- NULL
AdultUCI$"education-num" <- NULL
```

## Conceptual Hierarchy and Binning

```{r}
AdultUCI$age <- ordered(cut(AdultUCI$age, c(15,25,45,65,100)),
                              labels = c("Young", "Middle-aged", "Senior", "Old"))

AdultUCI$"hours-per-week" <- ordered(cut(AdultUCI$"hours-per-week",
                                             c(0,25,40,60,168)),
                                         labels = c("Part-time", "Full-time", "Over-time", "Workaholic"))

AdultUCI$"capital-gain" <- ordered(cut(AdultUCI$"capital-gain",
                                           c(-Inf,0,median(AdultUCI$"capital-gain"[AdultUCI$"capital-gain">0]),
                                             Inf)), labels = c("None", "Low", "High"))

AdultUCI$"capital-loss" <- ordered(cut(AdultUCI$"capital-loss",
                                           c(-Inf,0, median(AdultUCI$"capital-loss"[AdultUCI$"capital-loss">0]),
                                             Inf)), labels = c("None", "Low", "High"))

head(AdultUCI)
```

## Convert to transactions

```{r}
AdultTrans <- as(AdultUCI, "transactions")
```

## A Priori

```{r}
rules <- apriori(AdultTrans, parameter=list(supp = 0.5, conf = 0.9, minlen=2, maxlen= 10, target = "rules"), 
                 appearance=list(rhs = c("capital-gain=None"), default="lhs"), control=NULL)
inspect(rules)
```

```{r}
rules_a <- as(rules, "data.frame")
head(rules_a)
```

## Analysis of Rules

```{r}
imrules <- interestMeasure(rules, transactions = AdultTrans)
head(imrules)
```

## Removing redundant rules

```{r}
nrules <- rules[!is.redundant(rules)]
```

```{r}
arules::inspect(nrules)
```

## Showing the transactions that support the rules
In this example, we can see the transactions (trans) that support rules 1. 

```{r}
st <- supportingTransactions(nrules[1], AdultTrans)
trans <- unique(st@data@i)
length(trans)
print(c(length(trans)/length(AdultTrans), nrules[1]@quality$support))
```

Now we can see the transactions (trans) that support rules 1 and 2. 
As can be observed, the support for both rules is not the sum of the support of each rule. 

```{r}
st <- supportingTransactions(nrules[1:2], AdultTrans)
trans <- unique(st@data@i)
length(trans)
print(c(length(trans)/length(AdultTrans), nrules[1:2]@quality$support))
```

## Rules visualization

```{r}
options(repr.plot.width=10, repr.plot.height=5)
plot(rules)
```

```{r}
options(repr.plot.width=10, repr.plot.height=5)
plot(rules, method="paracoord", control=list(reorder=TRUE))
```

# Sequence Mining

```{r}
x <- read_baskets(con = system.file("misc", "zaki.txt", package = "arulesSequences"), info = c("sequenceID","eventID","SIZE"))
as(x, "data.frame")
```

```{r}
s1 <- cspade(x, parameter = list(support = 0.4), control = list(verbose = TRUE))
as(s1, "data.frame")
```

