---
title: An R Markdown document converted from "Rmd/tutorial/tut_r_course.ipynb"
output: html_document
---

#### Installation of R packages

```{r}
#install.packages("ISwR")
```

#### Package loading

```{r}
library(ISwR)
```

#### Variable definition and assignment

```{r}
weight <- 60
height = 1.75
subject <- "A"
healthy <- TRUE
```

#### Variable evaluation

```{r}
weight
```

#### Functions for type checking

```{r}
is.numeric(weight) # variable 
is.double(weight)
is.integer(weight)
is.character(subject)
```

#### Functions for variable conversion

```{r}
weight <- as.integer(weight)
is.integer(weight)
```

#### Computing the body mass index (BMI) from the weight and height

```{r}
#Body mass index (BMI)
bmi <- weight/height^2 
bmi 
```

#### Functions for string manipulation

```{r}
message <- sprintf("%.1f", bmi)
print(message)
```

#### Vector definition

```{r}
weight <- c(60, 72, 57, 90, 95, 72) 
height <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
subject <- c("A", "B", "C", "D", "E", "F")
```

#### Vector evaluation

```{r}
weight
height
subject
```

#### Creating a vector with a particular size

```{r}
vec <- rep(0, 10)
vec
```

#### Vector length

```{r}
length(weight)
```

#### Vector indexes: from one to the length of the vector

```{r}
weight[1]
weight[length(weight)]
```

#### Iteration: for loop
from one to the length of weight

```{r}
bmi <- 0
for (i in 1:length(weight)) {
  bmi[i] <- weight[i]/height[i]^2
}
```

evaluation of the bmi vector

```{r}
bmi
```

#### Iteration: while loop
run while i is below or equal to the length of weight

```{r}
bmi <- 0
i <- 1
while (i <= length(weight)) {
  bmi[i] <- weight[i]/height[i]^2
  i <- i + 1
}
```

```{r}
bmi
```

#### Remove a variable

```{r}
rm(bmi)
exists("bmi")
```

#### Right way of manipulating vectors: assigning at once

```{r}
bmi <- weight/height^2 
bmi 
```

#### Creating a function
name <- function(parameters) { body }

```{r}
compute_bmi <- function(weight, height) {
  bmi <- weight/height^2 
  return(bmi)
}
```

#### Using a function with scalars

```{r}

bmi <- compute_bmi(60, 1.75)
bmi
```

#### Using the same function with vectors

```{r}
bmi <- compute_bmi(weight, height)
bmi
```

#### Example of a function to compute the average
(iterating in all elements of the vector)

```{r}
average <- function(vec) {
    s <- 0
    n <- length(vec)
    for (x in vec) {
      s <- s + x  
    }
    return(s/n)
}
```

invoking the function

```{r}
avg_bmi <- average(bmi)
avg_bmi
```

#### Example of a function to compute the average
(manipulating vectors at once)

```{r}
average <- function(vec) {
    s <- sum(vec)
    n <- length(vec)
    return(s/n)
}
```

invoking the function

```{r}
avg_bmi <- average(bmi)
avg_bmi
```

#### Average function using mean function
Major statistical functions are available in R

```{r}
average <- function(vec) {
    return(mean(vec))
}
```

invoking the function

```{r}
avg_bmi <- average(bmi)
avg_bmi
```

#### Working with vectors with NA
Operations with NA lead to NA.

```{r}
x <- c(10, NA, 13)
y <- average(x)
y
```

#### addressing NA with na.rm=TRUE

```{r}
average <- function(vec) {
    return(mean(vec, na.rm=TRUE))
}
```

```{r}
x <- c(10, NA, 13)
y <- average(x)
y
```

#### Plotting graphics
scatter plots

```{r}
plot(height, weight)
```

#### Most functions contain many default parameters

```{r}
plot(height, weight, pch=2)
```

#### Default function arguments can be shown with args

```{r}
args(plot.default)
```

#### All functions in R that belongs to packages have help with examples

```{r}
?base::plot
```

#### Canvas for plotting is still active until a new plot

```{r}
plot(height, weight)
hh = c(1.65, 1.70, 1.75, 1.80, 1.85, 1.90)
lines(hh, 22.5 * hh^2)
```

#### Factors
Factors are used to handle categorical data.

```{r}
pain <- c(0,3,2,2,1)
fpain <- factor(pain,levels=0:3, ordered=TRUE)
fpain
```

#### Levels provide correspondence between numerical values and categorical labels

```{r}
levels(fpain) <- c("none","mild","medium","severe")
fpain
```

#### Convert height to factor
Levels: small, medium, high

#### coding setting element by element

```{r}
lev <- rep("", length(height))

for (i in 1:length(height)) {
  if (height[i] < 1.7)
    lev[i] <- "short"
  else if (height[i] < 1.9)
    lev[i] <- "medium"
  else 
    lev[i] <- "tall"
}
lev <- as.factor(lev)
lev
```

#### coding setting the vector at once
It uses the cut function.

```{r}
lev <- cut(height, breaks=c(0, 1.7, 1.9, .Machine$double.xmax), ordered=TRUE)
lev
levels(lev) <- c("short", "medium", "tall")
lev
```

#### Matrix
Matrices can be filled from vectors or data frames. 

```{r}
x <- 1:9
x
```

#### Converting a vector to matrix

```{r}
dim(x) <- c(3,3)
x
```

#### accessing elements from a matrix

```{r}
for (i in 1:nrow(x)) 
    for (j in 1:ncol(x))
        print(x[i,j])
    
```

#### Iterating and assigning values to each element

```{r}
y <- x
for (i in 1:nrow(y)) 
    for (j in 1:ncol(y))
        y[i,j] <- 3 * y[i, j]
    
y
```

#### Assigning the values of a matrix at once

```{r}
y <- 3*x
y
```

#### Converting a vector to a matrix by row

```{r}
x <- matrix(1:9,nrow=3,byrow=T)
x
```

#### transposing a matrix

```{r}
x <- t(x)
x
```

#### computing the determinant of a matrix

```{r}
det(x)
```

#### Lists
Lists are used to work with "objects"

```{r}
a <- c(5260,5470,5640,6180,6390,6515,6805,7515,7515,8230,8770)
b <- c(3910,4220,3885,5160,5645,4680,5265,5975,6790,6900,7335)

mybag <- list(a, b, 0, "a")
mybag
```

adding an element into a list

```{r}
n <- length(mybag)
mybag[[n+1]] <- "b"
mybag
```

#### List slicing

```{r}
slice <- mybag[1]
slice
is.list(slice)
```

#### Slicing is also a list

```{r}
slice <- mybag[c(1,3)]
slice
is.list(slice)
```

#### A list is also a vector

```{r}
#list is also a vector
is.vector(slice)
```

#### Member reference
It accesses the element

```{r}
h <- mybag[[1]]
h
```

An element can be evaluated. 
In this case, it is a vector.

```{r}
is.vector(h)
is.list(h)
```

#### Naming variables
They are properties on the list

```{r}
mybag <- list(x=a, y=b, const=0, lit="a")
mybag
```

#### Adding, accessing, and removing elements

```{r}
mybag$c <- mybag$x - mybag$y
mybag$const <- NULL
mybag$lit <- NULL
mybag
```

#### Data frames
Data frames (tables) provide support for structured data. 

```{r}
d <- data.frame(A=a, B=b)
head(d)
```

#### Adding a column in the data frame 

```{r}
d$c <- d$A + d$B
head(d)
```

```{r}
d$A <- NULL
head(d)
```

####  Reading csv file
There are many functions for reading CSV, Excel, and RData formats.

```{r}
wine = read.table(
    "http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data", 
                  header = TRUE, sep = ",")
  colnames(wine) <- c('Type', 'Alcohol', 'Malic', 'Ash', 
                      'Alcalinity', 'Magnesium', 'Phenols', 
                      'Flavanoids', 'Nonflavanoids',
                      'Proanthocyanins', 'Color', 'Hue', 
                      'Dilution', 'Proline')
head(wine)
```

####  saving in binary format 

```{r}
save(wine, file="wine.RData")
```

####  removing data frame from memory

```{r}
rm(wine)
```

####  load binary format

```{r}
load("wine.RData")
head(wine, 3)
```

####  exporting data.frame into csv file

```{r}
write.table(wine, file="wine.csv", row.names=FALSE, quote = FALSE, sep = ",")
```

####  filtering vectors

```{r}
a <- c(5260,5470,5640,6180,6390,6515,6805,7515,7515,8230,8770)
b <- c(3910,4220,3885,5160,5645,4680,5265,5975,6790,6900,7335)

# logical vector
bool <- (a > 7000)
bool
```

```{r}
# selecting elements from positions that are true
a[bool] 
```

```{r}
# filtering with logical expressions
b[a < 6000 | a > 7000]
```

```{r}
b[6000 <= a & a <= 7000]
```

####  filtering data frames

```{r}
data <- data.frame(a=a, b=b)
data$c <- data$a - data$b
head(data, nrow(data))
```

```{r}
head(data[data$a > 7000,])
```

```{r}
head(data[data$a > 7000,c(1,2)])
```

#### performance with matrix and data frames

```{r}
rheight <- rnorm(100000, 1.8, sd=0.2)
rweight <- rnorm(100000, 72, sd=15)
```

#### computing a entire column at once

```{r}
start_time <- Sys.time()
hw <- data.frame(height=rheight, weight=rweight)
hw$bmi <- hw$weight/hw$height^2
end_time <- Sys.time()
end_time - start_time
object.size(hw)
```

#### processing cell by cell

```{r}
start_time <- Sys.time()
hw <- data.frame(height=rheight, weight=rweight)
for (i in 1:nrow(hw)) {
  hw$bmi[i] <- hw$weight[i]/hw$height[i]^2
}
end_time <- Sys.time()
end_time - start_time
```

#### convert the entire column

```{r}
start_time <- Sys.time()
hw <- data.frame(height=rheight, weight=rweight)
hw <- as.matrix(hw)
hw <- cbind(hw, 0)
for (i in 1:nrow(hw)) {
  hw[i,3] <- hw[i,2]/hw[i,1]^2
}
end_time <- Sys.time()
end_time - start_time
```

#### apply family

apply functions can be applied for all rows or columns. 

The first character of the function name establishes the return type (s: simple, l: list).

```{r}
library(ISwR)
data(thuesen)
head(thuesen)
```

```{r}
#lapply returns a list
lapply(thuesen, mean, na.rm=T)
```

```{r}
#sapply returns a vector
sapply(thuesen, mean, na.rm=T)
```

```{r}
# apply - second parameter (1: by rows, 2: by columns)
m <- as.matrix(thuesen)
apply(m, 1, min, na.rm=TRUE)
apply(m, 2, min, na.rm=TRUE)
```

### sort and order

```{r}
library(ISwR)
data(thuesen)
head(thuesen)
```

```{r}
sort(thuesen$blood.glucose)
```

```{r}
order(thuesen$blood.glucose)
```

```{r}
o <- order(thuesen$blood.glucose)
sorted <- thuesen[o,]
head(sorted)
```

#### Pipelines
The operator $\%$>$\%$ creates a pipeline. 

The first parameter of the next invoked function receives the data from the pipeline. 

Library $dplyr$ contains a set of functions that support relational algebra operations.

```{r}
flight_data <- read.table(text = "Year Quarter Flights Delays
                     2016 1 11 6
                     2016 2 12 5
                     2016 3 13 3
                     2016 4 12 5
                     2017 1 10 4
                     2017 2 9 3
                     2017 3 11 4
                     2017 4 25 15
                     2018 1 14 3
                     2018 2 12 5
                     2018 3 13 3
                     2018 4 15 4",
                     header = TRUE,sep = "")  
head(flight_data)
```

```{r}
#install.packages("dplyr")
```

```{r}
library(dplyr)
result <- flight_data %>% 
   filter(Delays > 5) %>% 
   select(Year, Quarter, Flights)
head(result)
```

```{r}
library(dplyr)
result <- flight_data %>% 
   group_by(Year) %>% 
   summarize(mean = mean(Flights), sd = sd(Flights))
head(result)
```

```{r}
nrow(flight_data)
head(flight_data)
```

```{r}
#install.packages(reshape)
library(reshape)
result <- melt(flight_data[,c('Year', 'Quarter', 'Flights', 'Delays')], 
             id.vars = c(1,2))
nrow(result)
head(result[c(1:3,17:19), ])
```

#### merge

The function $merge$ can be used to join data frames. It can be used to produce inner, left, right, and outer joins. 

```{r}
stores <- data.frame(
  city = c("Rio de Janeiro", "Sao Paulo", "Paris", "New York", "Tokyo"),
  value = c(10, 12, 20, 25, 18))
head(stores)
```

```{r}
divisions <- data.frame(
  city = c("Rio de Janeiro", "Sao Paulo", "Paris", "New York", "Tokyo"),
  country = c("Brazil", "Brazil", "France", "US", "Japan"))
head(divisions)
```

```{r}
stdiv <- merge(stores, divisions, by.x="city", by.y="city")
head(stdiv)
```

```{r}
result <- stdiv %>% group_by(country) %>% 
   summarize(count = n(), amount = sum(value))
head(result)
```

#### statistical tests: t-test
There are many statistical tests in R.
One of the most used is the t-test. It checks if the mean of observations is not different from a theoretical value.

```{r}
weight <- c(60, 72, 57, 90, 95, 72) 
height <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
bmi <- weight/height^2 
```

```{r}
t.test(bmi, mu=22.5)
```

