# Exploratory analysis
A brief exploratory analysis example. 

### Iris datasets
The exploratory analysis is done using iris dataset.

```{r}
colors <- brewer.pal(4, 'Set1')
font <- theme(text = element_text(size=16))
```

```{r}
# DAL ToolBox
# version 1.0.77



library(daltoolbox)
library(ggplot2)
library(dplyr)
library(reshape)
library(RColorBrewer)
library(corrplot)
library(WVPlots)
library(GGally)
library(aplpack)

source("https://raw.githubusercontent.com/eogasawara/mylibrary/master/tutorial/graphics_extra.R")
```

## Part 1

```{r}
library(MASS)
data(iris)
```

```{r}
head(iris[c(1:3,51:53,101:103),])
```

### Data Summary
A preliminary analysis using the $Sepal.Length$ attribute. 

This should be done for all attributes. 

```{r}
sum <- summary(iris$Sepal.Length)
sum
```

```{r}
IQR <- sum["3rd Qu."]-sum["1st Qu."]
print(sprintf("IQR=%.1f", IQR))
```

## Part 2

```{r}
options(repr.plot.width=4, repr.plot.height=4)
```

### Histogram analysis

```{r}
grf <- plot_hist(iris %>% dplyr::select(Sepal.Length), 
          label_x = "Sepal Length", color=colors[1]) + font
plot(grf)
```

Grouping graphics

```{r}
grf1 <- plot_hist(iris %>% dplyr::select(Sepal.Length), 
                  label_x = "Sepal.Length", color=colors[1]) + font
grf2 <- plot_hist(iris %>% dplyr::select(Sepal.Width), 
                  label_x = "Sepal.Width", color=colors[1]) + font  
grf3 <- plot_hist(iris %>% dplyr::select(Petal.Length), 
                  label_x = "Petal.Length", color=colors[1]) + font 
grf4 <- plot_hist(iris %>% dplyr::select(Petal.Width), 
                  label_x = "Petal.Width", color=colors[1]) + font

load_library("gridExtra") 
options(repr.plot.width=16, repr.plot.height=4)
grid.arrange(grf1, grf2, grf3, grf4, ncol=4)
options(repr.plot.width=4, repr.plot.height=4)
```

### Density distribution

```{r}
grf1 <- plot_density(iris %>% dplyr::select(Sepal.Length), 
                  label_x = "Sepal.Length", color=colors[1]) + font
grf2 <- plot_density(iris %>% dplyr::select(Sepal.Width), 
                  label_x = "Sepal.Width", color=colors[1]) + font  
grf3 <- plot_density(iris %>% dplyr::select(Petal.Length), 
                  label_x = "Petal.Length", color=colors[1]) + font 
grf4 <- plot_density(iris %>% dplyr::select(Petal.Width), 
                  label_x = "Petal.Width", color=colors[1]) + font

options(repr.plot.width=16, repr.plot.height=4)
grid.arrange(grf1, grf2, grf3, grf4, ncol=4)
options(repr.plot.width=5, repr.plot.height=4)
```

### Box-plot analysis

```{r}
grf <- plot_boxplot(iris, colors=colors[1]) + font
plot(grf)
```

### Consider the classification problem targeting to predict the species

### Density distribution colored by the classifier

```{r}
grfA <- plot_density_class(iris %>% dplyr::select(Species, Sepal.Length), 
            class_label="Species", label_x = "Sepal.Length", color=colors[c(1:3)]) + font
grfB <- plot_density_class(iris %>% dplyr::select(Species, Sepal.Width), 
            class_label="Species", label_x = "Sepal.Width", color=colors[c(1:3)]) + font
grfC <- plot_density_class(iris %>% dplyr::select(Species, Petal.Length), 
            class_label="Species", label_x = "Petal.Length", color=colors[c(1:3)]) + font
grfD <- plot_density_class(iris %>% dplyr::select(Species, Petal.Width), 
            class_label="Species", label_x = "Petal.Width", color=colors[c(1:3)]) + font

options(repr.plot.width=8, repr.plot.height=8)
grid.arrange(grfA, grfB, grfC, grfD, ncol=2, nrow=2)
options(repr.plot.width=4, repr.plot.height=4)
```

### Box-plot analysis grouped by the classifier

```{r}
grfA <- plot_boxplot_class(iris %>% dplyr::select(Species, Sepal.Length), 
          class_label="Species", label_x = "Sepal.Length", color=colors[c(1:3)]) + font
grfB <- plot_boxplot_class(iris %>% dplyr::select(Species, Sepal.Width), 
          class_label="Species", label_x = "Sepal.Width", color=colors[c(1:3)]) + font
grfC <- plot_boxplot_class(iris %>% dplyr::select(Species, Petal.Length), 
          class_label="Species", label_x = "Petal.Length", color=colors[c(1:3)]) + font
grfD <- plot_boxplot_class(iris %>% dplyr::select(Species, Petal.Width), 
          class_label="Species", label_x = "Petal.Width", color=colors[c(1:3)]) + font

options(repr.plot.width=8, repr.plot.height=8)
grid.arrange(grfA, grfB, grfC, grfD, ncol=2, nrow=2)
options(repr.plot.width=5, repr.plot.height=4)
```

## Part 3

### Scatter plot

```{r}
grf <- plot_scatter(iris |> dplyr::select(x=Sepal.Length, value=Sepal.Width) |> mutate(variable = "iris"), 
                    label_x = "Sepal.Length", 
                    label_y = "Sepal.Width", colors=colors[1]) +
                    theme(legend.position = "none") + font
plot(grf)
```

```{r}
grf <- plot_scatter(iris |> dplyr::select(x = Sepal.Length, value = Sepal.Width, variable = Species), 
                    label_x = "Sepal.Length", label_y = "Sepal.Width", colors=colors[1:3]) + font

plot(grf)
```

### Correlation matrix

```{r}
grf <- plot_correlation(iris %>% 
                 dplyr::select(Sepal.Width, Sepal.Length, Petal.Width, Petal.Length))
options(repr.plot.width=8, repr.plot.height=8)
grf
options(repr.plot.width=4, repr.plot.height=4)
```

## Matrix dispersion

```{r}
grf <- plot_pair(data=iris, cnames=colnames(iris)[1:4], 
                 title="Iris", colors=colors[1])

options(repr.plot.width=10, repr.plot.height=10)
plot(grf)
options(repr.plot.width=4, repr.plot.height=4)
```

## Matrix dispersion by the classifier

```{r}
grf <- plot_pair(data=iris, cnames=colnames(iris)[1:4], 
                 clabel='Species', title="Iris", colors=colors[1:3])
options(repr.plot.width=10, repr.plot.height=10)
plot(grf)
options(repr.plot.width=4, repr.plot.height=4)
```

## Advanced matrix dispersion

```{r}
grf <- plot_pair_adv(data=iris, cnames=colnames(iris)[1:4], 
                     title="Iris", colors=colors[1])
options(repr.plot.width=10, repr.plot.height=10)
grf
options(repr.plot.width=4, repr.plot.height=4)
```

## Advanced matrix dispersion with the classifier

```{r}
grf <- plot_pair_adv(data=iris, cnames=colnames(iris)[1:4], 
                        title="Iris", clabel='Species', colors=colors[1:3])
options(repr.plot.width=10, repr.plot.height=10)
grf
options(repr.plot.width=4, repr.plot.height=4)
```

## Parallel coordinates

```{r}
grf <- ggparcoord(data = iris, columns = c(1:4), group=5) + 
    theme_bw(base_size = 10) + scale_color_manual(values=colors[1:3]) + font

options(repr.plot.width=10, repr.plot.height=5)
plot(grf)
options(repr.plot.width=4, repr.plot.height=4)
```

## Images

```{r}
mat <- as.matrix(iris[,1:4])
x <- (1:nrow(mat))
y <- (1:ncol(mat))

options(repr.plot.width=12, repr.plot.height=5)
image(x, y, mat, col = brewer.pal(11, 'Spectral'), axes = FALSE,  
      main = "Iris", xlab="sample", ylab="Attributes")
axis(2, at = seq(0, ncol(mat), by = 1))
axis(1, at = seq(0, nrow(mat), by = 10))
options(repr.plot.width=4, repr.plot.height=4)
```

## Chernoff faces

```{r}
set.seed(1)
sample_rows = sample(1:nrow(iris), 25)

isample = iris[sample_rows,]
labels = as.character(rownames(isample))
isample$Species <- NULL

options(repr.plot.width=12, repr.plot.height=12)
faces(isample, labels = labels, print.info=F, cex=1)
options(repr.plot.width=4, repr.plot.height=4)
```

## Chernoff faces with the classifier

```{r}
set.seed(1)
sample_rows = sample(1:nrow(iris), 25)

isample = iris[sample_rows,]
labels = as.character(isample$Species)
isample$Species <- NULL

options(repr.plot.width=12, repr.plot.height=12)
faces(isample, labels = labels, print.info=F, cex=1)
options(repr.plot.width=4, repr.plot.height=4)
```

